#PRACTICAL MACHINE LEARNING COURSE PROJECT

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Initial Set-up
The following libraries were used in this project.
```{r}
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(caret)
```
This seed was used for reproducibilty purposes
```{r}
set.seed(1414)
```

##Getting the Data

The training set can be found here:
```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
The testing set can be found here:
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" 
```
Load the data into R
```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

##Paritioning the data into two
Split the data into a training and testing dataset, with 60% of observations going into the training dataset
```{r}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

##Cleaning the data
Remove NearZeroVar variables from the training and testing datasets
```{r}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]
```
Remove the ID column so it doesn't interfere with algorithms
```{r}
myTraining <- myTraining[c(-1)]
```

clean variables with more than 60% NA
```{r}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
  if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
    for(j in 1:length(trainingV3)) {
      if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
        trainingV3 <- trainingV3[ , -j]
      }   
    } 
  }
}
```
Set back to the original variable name
```{r}
myTraining <- trainingV3
rm(trainingV3)
```
Transform the Testing and My testing datsets
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # remove the classe column
myTesting <- myTesting[clean1]         
testing <- testing[clean2]             

dim(myTesting)
dim(testing)
```
Coerce the data into the same type to ensure propoer functionality of machine learning algorithms
```{r}
for (i in 1:length(testing) ) {
  for(j in 1:length(myTraining)) {
    if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
      class(testing[j]) <- class(myTraining[i])
    }      
  }      
}

testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```
##Using Decision Trees

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
```
Predicting
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```
Using confusion matrix to test the results
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

##Using Random Forests
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```
Predicting
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```
Using confusion matrix to test the results
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```


##Conclusion
Random Forsests yielded better results than Decision Trees with a 99.9% accuracy as compared to 86.11%. The expected out of sample error is 100-99.9 = .01%
